---
title: "pvalue_integration"
format: html
---

# Load packages

```{r}
library(tidyverse)
library(fgsea)
library(msigdbr)
load(file='/home/projects/22140/exercise7_data.Rdata')

setwd('..')
source("./functions/edgington.R")
source("./functions/augment_data.R")
source("./functions/gene_analysis.R")
source("./functions/geneset_analysis.R")
source("./functions/fgsea_functions.R") # loads: create_stats, run_fgsea, sig_paths
```

# Load data

```{r, warning=FALSE}
# Change directory to parent to access the result folder
setwd('..')

# DESeq2
GSE125856_results <- read_csv("./results/GSE125856_results.csv")|> drop_na()
# GSE125856_results <- readRDS("./results/GSE125856_results.rds")
GSE162653_results <- read_csv("./results/GSE162653_results.csv")|> drop_na()
# GSE162653_results <- readRDS("./results/GSE162653_results.rds")
GSE165932_results <- read_csv("./results/GSE165932_results.csv")|> drop_na()
# GSE165932_results <- readRDS("./results/GSE165932_results.rds")


# DEXeq
# dexseq_results_GSE125856 <- readRDS("./results/dexseq_results_GSE125856.rds") |> as.data.frame() |> as_tibble() |> drop_na()
dexseq_results_GSE162653 <- readRDS("./results/dexseq_results_GSE162653.rds") |> as.data.frame() |> as_tibble() |> drop_na()
dexseq_results_GSE165932 <- readRDS("./results/dexseq_results_GSE165932.rds") |> as.data.frame() |> as_tibble() |> drop_na()
fgsea_DESeq2_BP_GSE162653 <- readRDS("./results/fgsea_DESeq2_BP_GSE162653.rds")
```

# Integrate p-values of GSE162653

```{r}
# GSE162653_results |> dplyr::select(gene_id)
```

```{r}
# augment125856 <- augment_data(GSE162653_results, dexseq_results_GSE125856)
augment162653 <- augment_data(GSE162653_results, dexseq_results_GSE162653)
augment165932 <- augment_data(GSE165932_results, dexseq_results_GSE165932)
augment162653 |> head()
```

```{r}

setwd("..")
augment162653 |> write_csv("results/03_pval_integration/augment162653-01.csv")
augment165932 |> write_csv("results/03_pval_integration/augment165932-01.csv")
```

## Gene-level analysis

```{r}
# sig_counts125856 <- gene_analysis(augment125856)
sig_counts162653 <- gene_analysis(augment162653)
sig_counts165932 <- gene_analysis(augment165932)
```

```{r}
sig_counts162653 |> head()
sig_counts165932 |> head()
```

### Uniquely significant genes

```{r}
unique165932 <- sum(augment165932$sig_joint & !augment165932$sig_deseq2 & !augment165932$sig_dexseq)
unique165932
```

## Gene-set-level analysis

```{r}
BP_df = msigdbr(species = "human", category = "C5", subcategory = "BP")
BP_list = split(BP_df$ensembl_gene, BP_df$gs_name)
MF_df = msigdbr(species = "human", category = "C5", subcategory = "MF")
MF_list = split(MF_df$ensembl_gene, MF_df$gs_name)
```

```{r}

# Results for dataset GSE162653
res162653 <- geneset_analysis(augment162653, BP_list, MF_list)
unique_joint_162653 <- res162653$unique_joint
fgsea_joint_162653  <- res162653$fgsea_joint

# Results for dataset GSE165932
res165932 <- geneset_analysis(augment165932, BP_list, MF_list)
unique_joint_165932 <- res165932$unique_joint
fgsea_joint_165932  <- res165932$fgsea_joint

```

Plots for Fishers integration analysis

```{r}
# 162653
augment162653 <- augment162653 |> 
  mutate(
    joint_only = sig_joint & !sig_deseq2 & !sig_dexseq,
    category = case_when(
      joint_only ~ "Joint-only",
      sig_deseq2 | sig_dexseq ~ "DESeq2 or DEXSeq",
      TRUE ~ "Not significant"
    )
  )

p1 <- ggplot(augment162653, aes(
  x = -log10(pvalue_deseq2),
  y = -log10(pvalue_dexseq),
  color = category
)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(
    values = c(
      "Not significant" = "grey80",
      "DESeq2 or DEXSeq" = "#1f78b4",
      "Joint-only" = "red"
    )
  ) +
  labs(
    title = "Genes Significant Only After Fisher Integration",
    x = "-log10(DESeq2 p-value)",
    y = "-log10(DEXSeq p-value)",
    color = "Gene Class"
  ) +
  theme_minimal(base_size = 14)

ggsave("../figures/Sig_genes_fischer_162653.png", p1, width = 10, height = 7, dpi = 300)

```

```{r}
# 165932
augment165932 <- augment165932 |> 
  mutate(
    joint_only = sig_joint & !sig_deseq2 & !sig_dexseq,
    category = case_when(
      joint_only ~ "Joint-only",
      sig_deseq2 | sig_dexseq ~ "DESeq2 or DEXSeq",
      TRUE ~ "Not significant"
    )
  )

p1 <- ggplot(augment165932, aes(
  x = -log10(pvalue_deseq2),
  y = -log10(pvalue_dexseq),
  color = category
)) +
  geom_point(alpha = 0.5, size = 1) +
  scale_color_manual(
    values = c(
      "Not significant" = "grey80",
      "DESeq2 or DEXSeq" = "#1f78b4",
      "Joint-only" = "red"
    )
  ) +
  labs(
    title = "Genes Significant Only After Fisher Integration",
    x = "-log10(DESeq2 p-value)",
    y = "-log10(DEXSeq p-value)",
    color = "Gene Class"
  ) +
  theme_minimal(base_size = 14)

ggsave("../figures/Sig_genes_fischer_165932.png", p1, width = 10, height = 7, dpi = 300)
```

The scatterplot shows that Fisher’s p-value integration identifies additional genes (red) that exhibit consistent but individually weak evidence in both DESeq2 and DEXSeq, demonstrating that multi-layer integration reveals biological signals missed by single-omics analyses.

```{r}
# 162653
joint_pathways_df_162653 <- fgsea_joint_162653 %>%
  filter(pathway %in% unique_joint_162653) %>%
  mutate(
    logpadj = -log10(padj)
  ) %>%
  arrange(padj)

top_joint_162653 <- joint_pathways_df_162653 %>% 
  slice_min(padj, n = 20)

p162653 <- ggplot(top_joint_162653,
       aes(x = logpadj,
           y = reorder(pathway, logpadj),
           size = NES,
           color = NES)) +
  
  geom_point(alpha = 0.75) +
  
  scale_color_gradient(low = "#fca5a5", high = "#b91c1c") +

  geom_vline(xintercept = -log10(0.05),
             linetype = "dashed", color = "grey40") +

  labs(
    title = "Top Joint-only Enriched Pathways (GSE162653)",
    x = "-log10(adjusted p-value)",
    y = "Pathway",
    size = "NES",
    color = "NES"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 9, lineheight = 1.1),
    plot.title  = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

p162653

ggsave("../figures/joint_pathways_GSE162653.png", p162653, 
       width = 10, height = 7, dpi = 300)

```

```{r}
# 165932
joint_pathways_df_165932 <- fgsea_joint_165932 %>%
  filter(pathway %in% unique_joint_165932) %>%
  mutate(
    logpadj = -log10(padj)
  ) %>%
  arrange(padj)

top_joint_165932 <- joint_pathways_df_165932 %>% 
  slice_min(padj, n = 20)

p165932 <- ggplot(top_joint_165932,
       aes(x = logpadj,
           y = reorder(pathway, logpadj),
           size = NES,
           color = NES)) +
  
  geom_point(alpha = 0.75) +
  
  scale_color_gradient(low = "#fca5a5", high = "#b91c1c") +

  geom_vline(xintercept = -log10(0.05),
             linetype = "dashed", color = "grey40") +

  labs(
    title = "Top Joint-only Enriched Pathways (GSE165932)",
    x = "-log10(adjusted p-value)",
    y = "Pathway",
    size = "NES",
    color = "NES"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 9, lineheight = 1.1),
    plot.title  = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

p165932

ggsave("../figures/joint_pathways_GSE165932.png", p165932, 
       width = 12, height = 7, dpi = 300)

```

```{r}
# Integrating p-values from both datasets using Edgington's method.
integrated_datasets <- edgington(augment162653,augment165932) |> write_csv("../results/03_pval_integration/integrated_datasets.csv")
```

```{r}
# Save significant genes
integrated_sig <- integrated_datasets %>% filter(sig_edgington == TRUE)

write_csv(integrated_sig,
          "../results/03_pval_integration/integrated_sig01.csv")

```

```{r}
integrated_gene_classes <- integrated_datasets %>%
  mutate(
    class = case_when(
      sig_edgington ~ "Integrated-significant",
      TRUE ~ "Not significant"
    )
  )
table(integrated_gene_classes$class)
```

```{r}
sig_162653 <- augment162653 %>% 
  filter(sig_joint == TRUE) %>% 
  pull(groupID)

sig_165932 <- augment165932 %>% 
  filter(sig_joint == TRUE) %>% 
  pull(groupID)

# Genes significant in at least one dataset before integration
sig_before <- union(sig_162653, sig_165932)

```

```{r}
sig_after <- integrated_datasets %>% 
  filter(sig_edgington == TRUE) %>% 
  pull(groupID)
```

```{r}
sig_only_after <- setdiff(sig_after, sig_before)
length(sig_only_after)
#93
```

FGSEA

```{r}
create_stats <- function(df, pvalue_col, id_col = "gene_id") {
s <- -log10(df[[pvalue_col]])
names(s) <- df[[id_col]]
# In case of duplicate gene IDs, keep the strongest evidence
s <- tapply(s, names(s), max)
sort(s, decreasing = TRUE)
}
```

```{r}
stats_int <- create_stats(integrated_datasets, "joint_pval_edgington", id_col = "groupID")

```

```{r}
fg_int_BP <- fgseaMultilevel(
  pathways = BP_list,
  stats     = stats_int,
  minSize   = 15,
  maxSize   = 500,
  scoreType = "std"
)

sig_int_BP <- fg_int_BP %>% filter(padj < 0.05)

nrow(sig_int_BP)
```

```{r}
write_csv(sig_int_BP,
          "../results/03_pval_integration/fgsea_integrated_BP_sig.csv")

```

```{r}

```

```{r}
sig_joint_162653 <- fgsea_joint_162653 %>% 
  filter(padj < 0.05) %>% 
  pull(pathway)

sig_joint_165932 <- fgsea_joint_165932 %>% 
  filter(padj < 0.05) %>% 
  pull(pathway)

# Pathways significant in each individual dataset
paths_162 <- sig_joint_162653      # already a character vector
paths_165 <- sig_joint_165932      # already a character vector

# Pathways significant after Edgington integration
paths_int <- sig_int_BP$pathway    # column from fgsea table

# Unique pathways only after integration
paths_edgington_only <- setdiff(paths_int, union(paths_162, paths_165))

length(paths_edgington_only)
#303


```

```{r}
top10_edg <- sig_int_BP %>% 
  filter(pathway %in% paths_edgington_only) %>% 
  arrange(padj) %>% 
  slice_head(n = 10)

top10_edg

```

```{r}
write_csv(top10_edg, "../results/03_pval_integration/top10_edgington_only_BP.csv")

```

plot

```{r}
top10_edg <- sig_int_BP %>%
  filter(pathway %in% paths_edgington_only) %>%
  arrange(padj) %>%
  slice_head(n = 10) %>%
  mutate(
    logpadj = -log10(padj),
  )
top10_edg

```

```{r}
library(ggplot2)

p <- ggplot(top10_edg,
            aes(x = logpadj,
                y = reorder(pathway, logpadj),
                size = abs(NES),
                color = NES)) +
  
  geom_point(alpha = 0.8) +
  
  scale_color_gradient2(
    low = "#4575b4",   # blue (down-regulated)
    mid = "grey80",
    high = "#d73027",  # red (up-regulated)
    midpoint = 0
  ) +
  
  labs(
    title = "Top 10 Edgington-only BP Pathways",
    x = "-log10(adjusted p-value)",
    y = "Pathway",
    size = "|NES|",
    color = "NES"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

p

```

```{r}
ggsave("../figures/top10_edg_bubbleplot.png", p,
       width = 10, height = 7, dpi = 300)

```
