---
title: "Virtual Pulldown 1"
format: html
editor: visual
---

## Load Libraries and database

```{r}
library(tidyverse)
library(igraph)
library(msigdbr)
library(fgsea)
library(ggraph)
library(biomaRt)


# Load PPI database (Lab 7)
load("/home/projects/22140/exercise7_ppi.Rdata")
setwd("..") 
```

```{r}
#seeds for virtual pulldown (Taken from ppi network)
seeds <- c(
  "ENSG00000004059",
  "ENSG00000149182",
  "ENSG00000023318",
  "ENSG00000170348",
  "ENSG00000129083",
  "ENSG00000095139",
  "ENSG00000100196"
)

```

```{r}
# Build PPI Graph
g <- graph_from_data_frame(ppi, directed = FALSE)

```

```{r}
# Creating Virtual Pulldown Function

filter_virtual_pulldown <- function(subGraph, parentGraph, cutoff) {

if (is.null(V(subGraph)$name) || is.null(V(parentGraph)$name)) {
stop("Both graphs require a 'name' vertex attribute.")
}

vn <- V(subGraph)$name

deg_full     <- degree(parentGraph, v = vn, mode = "all")
deg_internal <- degree(subGraph, mode = "all")

res <- data.frame(
node = vn,
deg_internal = deg_internal,
deg_full = deg_full,
frac_internal = ifelse(deg_full > 0, deg_internal / deg_full, NA_real_),
stringsAsFactors = FALSE
)

nodes_to_keep <- res$node[
!is.na(res$frac_internal) & res$frac_internal >= cutoff
]

filtered_graph <- induced_subgraph(subGraph, vids = V(subGraph)[name %in% nodes_to_keep])

V(filtered_graph)$frac_internal <-
res$frac_internal[match(V(filtered_graph)$name, res$node)]

filtered_graph
}

```

```{r}
# STEP 1: Virtual Pulldown

# Check which of your 7 seeds exist in the PPI network
valid_seeds <- intersect(seeds, V(g)$name)

if (length(valid_seeds) == 0) {
  stop("None of the seed genes are present in the PPI network `ppi`.")
}

seed_neighbors <- ego(
  graph = g,
  order = 1,
  nodes = V(g)[name %in% valid_seeds],
  mode = "all"
)

# Flatten the list of ego nodes
sub_nodes_names <- unique(unlist(lapply(seed_neighbors, as_ids)))

# Build the subgraph
g_sub <- induced_subgraph(g, vids = V(g)[name %in% sub_nodes_names])

# STEP 2: Filter virtual pulldown based on internal connectivity

g_filt <- filter_virtual_pulldown(
  subGraph = g_sub,
  parentGraph = g,
  cutoff = 0.5
)

```

For different cutoffs, no of genes are decreasing but not drastically. The most concerning thing is one of the seed nodes is not in the cluster and has low degree. To get what is going on, I am going to use Enrichr and see for different cutoffs how does the gene enrichment changes

### Getting gene names from Ensembl IDs for Enrichr (just a sanity check coz this is driving me crazy will delete later)

```{r}
# Get ENSG IDs from your filtered graph
ens_ids <- V(g_filt)$name

ens_ids <- unique(ens_ids)
length(ens_ids)  # how many IDs to map


# Use Ensembl Genes for human
mart <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

# Request mapping: ensembl_gene_id -> hgnc_symbol and external_gene_name
attrs <- c("ensembl_gene_id", "hgnc_symbol", "external_gene_name")
map_df <- getBM(attributes = attrs, filters = "ensembl_gene_id",
                values = ens_ids, mart = mart)

# Keep one row per ENSG (prefer hgnc_symbol then external_gene_name)
map_df <- map_df[!duplicated(map_df$ensembl_gene_id), ]
map_df$symbol <- ifelse(map_df$hgnc_symbol != "", map_df$hgnc_symbol, map_df$external_gene_name)
map_df <- map_df[,c("symbol")]



```

```{r}
# all the genes from virtual pulldown
 V(g_filt)$name

```

```{r}
# Highlight seed genes

library(ggraph)
library(tidygraph)

g_filt_tg <- as_tbl_graph(g_filt)

ggraph(g_filt_tg, layout = "fr") +
  geom_edge_link(alpha = 0.3) +
  geom_node_point(aes(color = name %in% seeds), size = 2) +
  geom_node_text(
    aes(label = ifelse(name %in% seeds, name, "")),
    repel = TRUE,
    size = 3
  ) +
  scale_color_manual(values = c("skyblue", "red")) +
  theme_void()


```

## Network Statistics

```{r}
# Average node degree
mean(degree(g_filt_tg))

```

```{r}
# transitivity
transitivity(g_filt_tg)
```

```{r}
diameter(g_filt_tg)
```

```{r}
 # clustering the results
cl <- cluster_louvain(g_filt, weights = NULL, resolution = 0.4)
 
```

```{r}
# Enrichment
cluster_vec <- membership(cl)
V(g_filt)$cluster <- factor(cluster_vec)
n_clusters <- sort(table(V(g_filt)$cluster), decreasing = TRUE)
n_clusters
```

```{r}
clusters_df <- data.frame(
gene_id = names(cluster_vec),
cluster = as.integer(cluster_vec),
stringsAsFactors = FALSE
)
```

̥
